Cognate beginnings to bilingual lexical growth

### 2023-05-23 manuscript/manuscript.qmd
Add AoA estimator

tar_load(c(bvq_data, childes, items, participants, responses))
tar_load(c(starts_with("model_"), model_loos))
tar_load(c(posterior_draws, marginal_effects_epreds))

@@

@fig-hypothesis provides a graphic, theoretical illustration of this account. We assume the case of a Catalan-Spanish bilingual child who is exposed 65% of the time to Catalan, and 35% of the time to Spanish. On the top panels, we illustrate the cumulative sum of learning instances that a child accumulates across ages for the Catalan word-form (/\textipa{gat}/ or /\textipa{gos}/) and its Spanish translation (/\textipa{"pe.ro}/ or /\textipa{"ga.to}/). We assume an arbitrary, theoretical threshold for the learning instances at which the child is considered to have acquired a word (solid grey line). Points and triangles show the age-of-acquisition at which the amount of learning instances reaches the threshold. We show the predicted age-of-acquisition under two hypothesis: (1) a cognateness facilitation effect that lowers the acquisition threshold after one of the word-forms (in this case Catalan /\textipa{gat}/) has been acquired, and (2) no cognateness facilitation, for which the acquisition of the second word-form occurs unconditionally of the acquisition of the first. On the bottom, we show the predicted probability of acquisition for the Catalan and Spanish words under each hypothesis, as it would be expected to observe in a vocabulary questionnaire like the one in the present study. We assume that such probability starts at 0% at birth, reaches 50% (and grows the fastest) at the age-of-acquisition, and reaches 100% at some points in later ages.

@@

@@

diff_responses <- responses |>
    distinct(id, age) |>
    left_join(distinct(participants, id, time_stamp, age),
              by = c("id", "age")) |> 
    group_by(id) |> 
    arrange(id, time_stamp) |> 
    mutate(diff_lag = time_stamp-dplyr::lag(time_stamp, default = NA)) |> 
    summarise(diff_min = min(diff_lag, na.rm = TRUE),
              diff_max = max(diff_lag, na.rm = TRUE),
              .groups = "drop") |> 
    filter(!(diff_min %in% c(-Inf, Inf))) |>
    summarise(diff_min = min(diff_min, na.rm = TRUE),
              diff_max = max(diff_max, na.rm = TRUE))

@@

```{r tbl-participants}
#| label: tbl-participants
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan."
#| tbl-note: "Information about language degree of exposure (DoE) was provided by families before filling the questionnaire. In this table, a 100\\% indicates that the participant was exclusively exposed to Catalan, and never to Spanish. A 0\\% indicates that the participant was not exposed to Catalan ever, and most of the time to Spanish. A 50\\% indicates that the participant was exposed to Catalan and Spanish approximately half of the time each. The 100\\%, 0\\%, and 50\\% would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively. For illustration purposes this table, DoEs were binned into 25\\%-wide bins, and ages (in months) binned into 4 month-wide bins. Hyphens indicate that no participants from that specific combination of age and DoE filled the questionnaire."
participants |>
    mutate_at(vars(starts_with("doe_")), 
              (\(x) floor(x * 10) / 10)) |> 
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = cut(
            doe_catalan,
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%",
                       "25-50%", 
                       "50-75%",
                       "75-100%"),
            include.lowest = TRUE
        ) |>
            as.character()
    ) |>
    count(age, doe_catalan) |>
    arrange(desc(doe_catalan)) |>
    pivot_wider(names_from = age, values_from = n) |>
    gt() |>
    summary_rows(
        columns = 2:7,
        fns = list("Total" = ~ sum(., na.rm = TRUE)),
        formatter = fmt_integer
    ) |>
    sub_missing(everything(), everything(), missing_text = "--") |>
    cols_label(doe_catalan = "Catalan exposure") |>
    tab_spanner("Age (months)", 2:8) |>
    tab_style(cell_text(align = "left"),
              list(cells_body(doe_catalan),
                   cells_column_labels(doe_catalan)))
```

@@

n_items_nouns <- n_items |> 
    filter(class=="Noun")

@@

n_items_lemma <- n_items_freq |> 
    filter(n_lemmas == 1)

n_items_multi <- n_items_lemma |> 
    filter(!is_multiword) |> 
    filter(include, !is.na(wordbank_lemma)) |> 
    drop_na(xsampa) |> 
    distinct(te, language)

n_items_duplicated <- n_items_multi |> 
    add_count(te, name = "n_te") |>
    filter(n_te == 2)

@@

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, mixed-effects ordinal regression model (`r dim(model_fit_0$fit)[2]` sampling chains with `r format(dim(model_fit_0$fit)[1], big.mark = ",")` iterations each). This model allowed us to estimate both item and participant word-acquisition trajectories, as a function of different fixed effects from $\text{Age}$ (number of months elapsed between participants' birth date and questionnaire completion), $\text{Length}$ (number of phonemes in the X-SAMPA phonological transcription of the word-form), $\text{Exposure}$ (composite measure of lexical frequency and language degree of exposure), and $\text{Cognateness}$ (a measure of cognateness, defined as the phonological similarity between translation equivalents). Phonological transcriptions were manually generated. Catalan word-forms were transcribed to central Catalan phonology, and Spanish translations were transcribed to  Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor. Participant- and item-level random intercepts and slopes were included when appropriate, according to the structure of the data [@barr2013random]. We specified a weakly informative prior around the parameters of the model (see Appendix C for details).

$$
\begin{aligned}
&\textbf{Response}~k~\textbf{to word}~i~ \textbf{by participant}~j \\
\text{Response}_{ij} &\sim \text{Cumulative logit}(\theta_{k_{ij}}), \\ &\text{where}~k \in \{\text{No, Understands, Understands and Says}\}\\
&\textbf{Distribution parameters} \\
\theta_{k_{ij}} &= (\beta_{0_{k}} + u_{0_{k_{i}}} + w_{0_{k_{j}}}) + (\beta_{1} + u_{1_{i}} + w_{1_{j}}) \cdot \text{Age}_{i} + \\
& (\beta_{2} + u_{2_{i}} + w_{2_{j}}) \cdot \text{Length}_{ij} + (\beta_{3} + u_{3_{i}} + w_{3_{j}}) \cdot \text{Exposure}_{ij} + & \\
& (\beta_{4} + u_{4_{i}}) \cdot \text{Cognateness}_{i} + (\beta_{5} + u_{5_{i}}) \cdot (\text{Exposure}_{ij} \times \text{Cognateness}_{i}) & \\
u_{0-5_{i}} &\sim \mathcal{N}(0, \sigma_{u_{0-8_{i}}}); ~w_{0-3_{j}} \sim \mathcal{N}(0, \sigma_{w_{0-3_{j}}}) \\
&\textbf{Prior} \\
\beta_{0_{k}} &\sim \mathcal{N}(-0.25, 0.5); ~\beta_{1-5} \sim \mathcal{N}(0, 1) \\
\sigma_{u_{0-5}, w_{0-3}} &\sim \mathcal{N_{+}}(1, 0.25); ~\rho_{u_{0-8}, w_{0-3}} \sim \text{LKJcorr}(2)
\end{aligned}
$$

We built a base model $\mathcal{M}_{0}$ that only included the intercept and the main effects of $\text{Age}$, $\text{Length}$, and $\text{Exposure}$ as predictors, and compared it against increasingly more complex models. $\mathcal{M}_{1}$ extended the base model by adding the main effect of $\text{Cognateness}$, and $\mathcal{M}_{2}$ added the two-way interaction $\text{Exposure} \times \text{Cognateness}$. We compared the predictive performance of the models using Bayesian leave-one-out cross-validation [LOO-CV, @vehtari2017practical]^[Due to computational constraints, we performed LOO-CV based on a random sub-sample of `r format(dim(model_log_liks$model_fit_0)[1], big.mark = ",")` samples of the posterior distribution of the fixed effects [@magnusson2020leaveoneout]]. LOO-CV involves computing the likelihood of each data point, given a model  fitted excluding such data point. The likelihoods generated by the model for each of the data points are summarised together in a measure of predictive performance: the expected log-predicted density (ELPD). This measure penalises the complexity of the model ELPD. ELPD values closer to zero indicate better predictive accuracy. We compared the ELPD of each model, and selected for interpretation the one showing the best performance (see Appendix C for more details about the models).

We assessed the practical relevance of the estimated regression coefficients of the selected model following @kruschke2018bayesian. First, we specified a region of practical equivalence (ROPE) from `r rope_interval["lower"]` to +`r rope_interval["upper"]`, in the probability scale. This region indicates the range of values that we considered equivalent to zero. The degree of overlap between the posterior distribution of a regression coefficient with the  can be interpreted as support for the true parameter of the coefficient being zero or equivalent. We then computed the 95% highest density interval (HDI) of the posterior distribution of each regression coefficient. This interval contains the true value of this coefficient with 95% probability, given the data. Finally, we calculated the proportion of posterior samples in the 95% HDI that fell into the ROPE, which indicates the probability that the true value of the regression coefficient falls into the ROPE, and should therefore be considered equivalent to zero. For example, a 90% overlap between the HDI of a coefficient's posterior distribution and the  indicates that, given our data, there is a 90% probability that the true value of the coefficient falls within the ROPE, and is therefore equivalent to zero.

Data processing and visualisation was done in R [@rcoreteam2013language, version `r r_version`] using the Tidyverse family of packages [@wickham2019welcome]. Bayesian modelling was done using the `brms` [@burkner2017brms; @carpenter2017stan], `loo` [@vehtari2022loo], `tidybayes` [@kay2021tidybayes] and `marginaleffects` [@arel-bundock2022marginaleffects] R packages.

@@

```{r results-values}
#| label: results-values
#| echo: false
rhats_df <- map_df(model_rhats, bind_rows, .id = "model") |> 
    mutate(model = paste0("Model ", str_extract(model, "[0-9]"))) |> 
    pivot_longer(-model, 
                 names_to = ".variable",
                 values_to = ".rhat") |>
    filter(str_detect(.variable, "b_|sd_")) |> 
    drop_na()

loo_diff <- loo_compare(model_loos)[, 1]
loo_diff_se <- loo_compare(model_loos)[, 2]

post_draws_list <- posterior_draws$summary |> 
    mutate(across(.median:.lower,
                  \(x) ifelse(grepl("Intercept", .type),
                              plogis(x), x/4))) |> 
    split(posterior_draws$summary$.variable)
```

All models showed adequate convergence diagnostics and little evidence of multicollinearity between predictors (see Appendix D for details). $\mathcal{M}_{2}$, which included the $\text{Exposure} \times \text{Cognateness}$ two-way interaction, showed the best predictive performance with a difference in expected log-predicted density (ELPD) compared to the second best-performing model ($\mathcal{M}_{1}$) of `r round(loo_diff[2], 2)` $\pm$ `r round(loo_diff_se[2], 2)`. $\mathcal{M_1}$ performed substantially better than $\mathcal{M_0}$, with a difference of `r round(loo_diff[3], 2)` $\pm$ `r round(loo_diff_se[3], 2)` This points to the effect of $\text{Cognateness}$ being strongly modulated by $\text{Exposure}$, and that this interaction remains relatively stable across ages.

```{r tbl-results-loos}
#| label: tbl-loos
#| echo: false
#| message: false
#| warning: false
#| eval: true
#| tbl-cap: "Bayesian leave-one-out cross validation (LOO-CV)."
#| tbl-note: "ELPD (*SD*): expected log-predictive density and standard error. Diff. (*SD*): diference in ELPD between each model and the model with the best predictive accuracy (Model 3) and standard error of the difference. Pareto-k estimates of all models were acceptable (k < 0.5)."
model_loos |>
    loo_compare() |>
    as.data.frame() |>
    rownames_to_column("model") |>
    mutate(model = paste0("Model ", str_remove(model, "model_fit_"))) |>
    select(model, elpd_loo, se_elpd_loo, elpd_diff, se_diff) |> 
    gt(rowname_col = "model") |>
    fmt_number(2:5, decimals = 2, sep_mark = ",") |>
    cols_merge(c(elpd_loo, se_elpd_loo), pattern = "{1} ({2})") |> 
    cols_merge(c(elpd_diff, se_diff), pattern = "{1} ({2})") |> 
    cols_label(
        model = "Model",
        elpd_loo = "ELPD (SD)",
        elpd_diff = "Diff. (SD)"
    ) |>
    tab_style(cell_text(weight = "bold", style = "italic"),
              cells_column_labels()) |>
    tab_style(cell_text(align = "left"),
              cells_title(groups = "subtitle"))
```

@tbl-coefs shows the summary of the posterior distribution of the fixed regression coefficients of $\mathcal{M}_{2}$, and their degree of overlap with the ROPE. For interpretability, we report the estimated regression coefficients transformed to the probability scale. The resulting values correspond to the maximum difference in probability of acquisition (*Comprehension* or *Comprehension and Production*) that corresponds to a one standard deviation change in each predictor^[The logit and probability scales relate non-linearly. This means that one logit difference is not necessarily translated to a unique value in the probability scale. For example, the probability of acquisition of a given word might increase in 5% when age increases from 22 to 23 months, the probability of acquisition of the same word might only increase in 0.2% when age increases from 30 to 32 months. The linear growth of the probability of acquisition differs along the logistic curve, and therefore deciding the age point at which to report the estimates of the regression coefficients in the probability scale is not trivial. Following @gelman2020regression, we report the maximum value of such coefficient, which corresponds to the linear growth (i.e. derivative) of the logistic curve at the age at which most participants were acquiring a given word. This value can be approximated by dividing the coefficient in the logit scale by four: $\hat{\beta_j}/4$, where $\hat{\beta_j}$ is the estimated mean of the posterior distribution of coefficient $j$.].

```{r tbl-coefs}
#| label: tbl-coefs
#| tbl-cap: "Posterior distribution of regression coefficients Model 3."
#| tbl-note: "Median: median of the posterior distribution in the probability scale. 95\\% HDI: 95\\% highest density interval of the distribution; narrowest range of values that contains 95\\% of the distribution, and therefore is 95\\% likely to contain the true value, given the data. On top of each distribution, we indicate the proportion of posterior samples in the 95\\% HDI that fall into the region of practical equivalence (ROPE). This proportion indicates the probability of the true value of the coefficient being equivalent to zero. Lower values indicate that the true value is unlikely to be zero of equivalent to zero."
#| echo: false
#| message: false
#| warning: false
#| out-width: 80%
# summarise posterior draws
intercept_label <- "Intercepts (at 22 months)"

posterior_draws$summary |>
    select(.variable_name, .median, .lower, .upper, .rope_overlap, .type) |>
    mutate(across(.median:.upper,
                  \(x) ifelse(.type==intercept_label, plogis(x), x/4))) |> 
    gt(groupname_col = ".type", 
       rowname_col = ".variable_name") |>
    fmt_percent(.rope_overlap, decimals = 3) |> 
    fmt_number(c(.median:.upper), decimals = 3) |>
    cols_merge(c(.lower, .upper),
               pattern = "[{1}, {2}]") |>
    cols_label(
        .variable_name = "Parameter",
        .median = "Median",
        .lower = md("95% HDI"),
        .rope_overlap = md("ROPE prob.")
    ) |>
    tab_style(cell_text(weight = "bold"),
              cells_column_labels(columns = 1:5)) |>
    tab_style(cell_text(align = "left"),
              cells_title(groups = "subtitle")) |> 
    tab_style(cell_text(size = "small"), 
              list(cells_body(),
                   cells_row_groups(),
                   cells_stub())) |> 
    tab_style(cell_text(style = "italic"),
              cells_row_groups())
```

@@

```{r fig-marginal}
#| label: fig-marginal
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.5
#| out-width: 100%
#| eval: true
#| fig-cap: "Posterior marginal effects of Model 3. Thin lines correspond to 50 mean predictions. Thicker lines indicate the median of those predictions. Different colours indicate different levels of cognateness (phonological similarity). Predictions are presented separately for different degrees of word exposure index: little exposure to the word, mean exposure, and high exposure. Predictions for Comprehension are show on top and predictions for Comprehension and Production are shown on the bottom. In-sample predictions lie inside the grey rectangles."
clrs_reorder <- c("#58508d", "#003f5c", "#122A36", "#ffa600", "#ff6361", "#bc5090")

marginal_effects_epreds |> 
    mutate(age = rescale_variable(age_std, 
                                  mean(responses$age), 
                                  sd(responses$age)),
           exposure = factor(exposure_std,
                             levels = unique(exposure_std),
                             labels = c("Low exposure (-1 SD)",
                                        "Mean exposure",
                                        "High exposure (+1 SD)"),
                             ordered = TRUE),
           lv = factor(lv_std,
                       levels = unique(lv_std),
                       labels = paste0(c(0, 50, 100), "%"),
                       ordered = TRUE),
           .category = ifelse(.category=="Understands",
                              "Comprehension",
                              "Production")) |> 
    ggplot(aes(age, .value, colour = lv, fill = lv)) +
    facet_grid(.category~exposure) +
    annotate(geom = "rect",
             xmin = min(responses$age),
             xmax = max(responses$age),
             ymin = -Inf,
             ymax = Inf,
             fill = "grey95",
             colour = NA) +
    annotate(geom = "text",
             x = mean(responses$age),
             y = 0,
             hjust = -0.1,
             vjust = 0,
             label = "Observed age range",
             size = 2) +
    geom_vline(xintercept = mean(responses$age), 
               linetype = "dotted") +
    geom_hline(yintercept = 0.5,
               linetype = "dotted") +
    geom_line(aes(group = interaction(lv, .draw)),
              linewidth = 0.5,
              alpha = 1/10) +
    stat_summary(fun = median, geom = "line", linewidth = 0.75) +
    labs(x = "Age (months)",
         y = "Probability of acquisition",
         linetype = "Cognateness (phonological similarity)",
         colour = "Cognateness (phonological similarity)",
         fill = "Cognateness (phonological similarity)") +
    guides(colour = guide_legend(override.aes = list(linewidth = 2))) +
    scale_colour_manual(values = clrs_reorder[c(4, 5, 1)]) +
    scale_fill_manual(values = clrs_reorder[c(4, 5, 1)]) +
    scale_y_continuous(breaks = seq(0, 1, 0.25),
                       limits = c(0, 1)) +
    scale_x_continuous(breaks = seq(0, 50, 5)) +
    theme(legend.position = "top",
          legend.key.size = unit(1, "cm"),
          legend.key.height = unit(1, "cm"),
          axis.text.x = element_text(size = 9),
          panel.grid = element_blank(),
          panel.border = element_rect(fill = NA, 
                                      colour = "black", 
                                      linewidth = 0.75))
```

@@

E_e <- bvq_data$pool |>
    mutate(item = gsub("cat_|spa_", "", item),
           xsampa = gsub("\\.", "", xsampa)) |>
    filter(te %in% items$te, 
           item %in% items$item) |>
    select(te, language, xsampa) |> 
    pivot_wider(names_from = language,
                values_from = xsampa, 
                id_cols = te,
                names_repair = janitor::make_clean_names) |>
    filter(grepl("E", catalan), 
           grepl("e", spanish)) |>
    left_join(distinct(items, te, lv),
              by = "te")
```

>>>>

tar_load(c(bvq_data, childes, items, participants, responses))
tar_load(c(model_fit, model_summary, model_epreds, model_convergence))

@@

@fig-hypotheses provides a graphic, theoretical illustration of this account. We assume the case of a Catalan-Spanish bilingual child who is exposed 65% of the time to Catalan, and 35% of the time to Spanish. On the top panels, we illustrate the cumulative sum of learning instances that a child accumulates across ages for the Catalan word-form (/\textipa{gat}/ or /\textipa{gos}/) and its Spanish translation (/\textipa{"pe.ro}/ or /\textipa{"ga.to}/). We assume an arbitrary, theoretical threshold for the learning instances at which the child is considered to have acquired a word (solid grey line). Points and triangles show the age-of-acquisition at which the amount of learning instances reaches the threshold. We show the predicted age-of-acquisition under two hypothesis: (1) a cognateness facilitation effect that lowers the acquisition threshold after one of the word-forms (in this case Catalan /\textipa{gat}/) has been acquired, and (2) no cognateness facilitation, for which the acquisition of the second word-form occurs unconditionally of the acquisition of the first. On the bottom, we show the predicted probability of acquisition for the Catalan and Spanish words under each hypothesis, as it would be expected to observe in a vocabulary questionnaire like the one in the present study. We assume that such probability starts at 0% at birth, reaches 50% (and grows the fastest) at the age-of-acquisition, and reaches 100% at some points in later ages.

@@

For each word included in the questionnaire, we manually generated a broad phonological transcription in X-SAMPA format. Catalan word-forms were transcribed to Central Catalan phonology, and Spanish translations were transcribed to Castilian Spanish phonology. For each translation equivalent, we calculated the phonological similarity between the Catalan and the Spanish word-form using the same Levenshtein similarity measure used in @floccia2018introduction.

@@

suppressWarnings({
    diff_responses <- responses |>
        distinct(id, age) |>
        left_join(distinct(participants, id, time_stamp, age),
                  by = c("id", "age")) |> 
        group_by(id) |> 
        arrange(id, time_stamp) |> 
        mutate(diff_lag = time_stamp-dplyr::lag(time_stamp, default = NA)) |> 
        summarise(diff_min = min(diff_lag, na.rm = TRUE),
                  diff_max = max(diff_lag, na.rm = TRUE),
                  .groups = "drop") |> 
        dplyr::filter(!(diff_min %in% c(-Inf, Inf))) |>
        summarise(diff_min = min(diff_min, na.rm = TRUE),
                  diff_max = max(diff_max, na.rm = TRUE))
})

@@

```{r tbl-participants}
#| label: tbl-participants
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan."
#| tbl-note: "Information about language degree of exposure (DoE) was provided by families before filling the questionnaire. In this table, a 100\\% indicates that the participant was exclusively exposed to Catalan, and never to Spanish. A 0\\% indicates that the participant was not exposed to Catalan ever, and most of the time to Spanish. A 50\\% indicates that the participant was exposed to Catalan and Spanish approximately half of the time each. The 100\\%, 0\\%, and 50\\% would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively. For illustration purposes this table, DoEs were binned into 25\\%-wide bins, and ages (in months) binned into 4 month-wide bins. Hyphens indicate that no participants from that specific combination of age and DoE filled the questionnaire."
participants |>
    mutate_at(vars(starts_with("doe_")), 
              (\(x) floor(x * 10) / 10)) |> 
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = cut(
            doe_catalan,
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%",
                       "25-50%", 
                       "50-75%",
                       "75-100%"),
            include.lowest = TRUE
        ) |>
            as.character()
    ) |>
    count(age, doe_catalan) |>
    arrange(desc(doe_catalan)) |>
    pivot_wider(names_from = age, values_from = n) |>
    gt() |>
    sub_missing(everything(), everything(), missing_text = "--") |>
    cols_label(doe_catalan = "Catalan exposure") |>
    tab_spanner("Age (months)", 2:7) |>
    tab_style(cell_text(align = "left"),
              list(cells_body(doe_catalan),
                   cells_column_labels(doe_catalan)))
```

@@

n_items_nouns <- n_items |> 
    dplyr::filter(class=="Noun")

@@

n_items_lemma <- n_items_freq |> 
    dplyr::filter(n_lemmas == 1)

n_items_multi <- n_items_lemma |> 
    dplyr::filter(!is_multiword) |> 
    dplyr::filter(include, !is.na(wordbank_lemma)) |> 
    drop_na(xsampa) |> 
    distinct(te, language)

n_items_duplicated <- n_items_multi |> 
    add_count(te, name = "n_te") |>
    dplyr::filter(n_te == 2)

@@

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, mixed-effects ordinal regression model. This model allowed us to estimate both item and participant word-acquisition trajectories, while estimating the effect of our variables of interest. We added $\text{Age}$ (number of months elapsed between participants' birth date and questionnaire completion), $\text{Length}$ (number of phonemes in the X-SAMPA phonological transcription of the word-form), $\text{Exposure}$ (composite measure of lexical frequency and language degree of exposure), and $\text{Cognateness}$ (a measure of cognateness, defined as the phonological similarity between translation equivalents) as population-level effects, together with their two- and three-way interactions. Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor. Participant-level and item-level random intercepts and slopes were included where appropriate, according to the structure of the data [@barr2013random; see @eq-linear]. We specified a weakly informative prior around the parameters of the model (see Appendix C for details).

$$
\begin{aligned}
&\textbf{Response}~(k)~\textbf{to word}~i~ \textbf{by participant}~j \\
\text{Response}_{ij} &\sim \text{Cumulative logit}(\theta_{k_{ij}}), \\ &\text{where}~k \in \{\text{No, Understands, Understands and Says}\}\\
&\textbf{Distribution parameters} \\
\theta_{k_{ij}} &= (\beta_{0_{k}} + u_{0_{i_{k}}} + w_{0_{j_{k}}}) + (\beta_{1} + u_{1_{i}} + w_{1_{j}}) \cdot \text{Age}_{i} + \\
&(\beta_{2} + u_{2_{i}} + w_{2_{j}}) \cdot \text{Length}_{ij} + 
(\beta_{3} + u_{3_{i}} + w_{3_{j}}) \cdot \text{Exposure}_{ij} + & \\
& (\beta_{4} + u_{4_{i}}) \cdot \text{Cognateness}_{ij} + (\beta_{5} + u_{5_{i}} + w_{3_{j}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij}) + & \\
& (\beta_{6} + u_{6_{i}}) \cdot (\text{Age}_{i} \times \text{Cognateness}_{ij}) + & \\
& (\beta_{7} + u_{7_{i}}) \cdot (\text{Exposure}_{ij} \times \text{Cognateness}_{ij}) & \\
& (\beta_{8} + u_{8_{i}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij} \times \text{Cognateness}_{ij}) \\
&\textbf{Prior} \\
\beta_{0_{k}} &\sim \mathcal{N}(-0.25, 0.5); ~\beta_{1-5} \sim \mathcal{N}(0, 1) \\
\sigma_{u_{0-8}, w_{0-3}} &\sim \mathcal{N_{+}}(1, 0.25); ~\rho_{u_{0-8}, w_{0-3}} \sim \text{LKJcorr}(2)
\end{aligned}
$$ {#eq-linear}

We assessed the practical relevance of the estimated regression coefficients of the model model following @kruschke2018bayesian. First, we specified a region of practical equivalence (ROPE) from `r rope_interval["lower"]` to +`r rope_interval["upper"]`, in the probability scale. This region indicates the range of values that we considered equivalent to zero. The degree of overlap between the posterior distribution of a regression coefficient with the  can be interpreted as support for the true parameter of the coefficient being zero or equivalent. We then computed the 95% highest density interval (HDI) of the posterior distribution of each regression coefficient. This interval contains the true value of this coefficient with 95% probability, given the data. Finally, we calculated the proportion of posterior samples in the 95% HDI that fell into the ROPE, which indicates the probability that the true value of the regression coefficient falls into the ROPE, and should therefore be considered equivalent to zero. For example, a 90% overlap between the HDI of a coefficient's posterior distribution and the  indicates that, given our data, there is a 90% probability that the true value of the coefficient falls within the ROPE, and is therefore equivalent to zero.

Data processing and visualisation was done in R [@rcoreteam2013language, version `r r_version`] using the Tidyverse family of packages [@wickham2019welcome]. We implemented the model using was done using the `brms` [@burkner2017brms], a R interface to the Stan probabilistic language [@carpenter2017stan]. We ran `r nchains(model_fit)` MCMC  chains with `r format(niterations(model_fit), big.mark = ",")` iterations each and an additional `r format(niterations(model_fit), big.mark = ",")` warm-up iterations per chain. Model posterior draws and predictions were handled using the `tidybayes` [@kay2021tidybayes] and `marginaleffects` [@arel-bundock2022marginaleffects] R packages.

@@

```{r results-values}
#| label: results-values
#| echo: false
rhats_df <- model_convergence |> 
    dplyr::filter(str_detect(.variable, "b_|sd_")) |> 
    drop_na()

post_draws_list <- model_summary |> 
    mutate(across(.median:.lower,
                  \(x) ifelse(grepl("Intercept", .type),
                              plogis(x), x/4))) |> 
    split(model_summary$.variable)
```

All parameters in the model showed adequate MCMC convergence diagnostics and little evidence of multicollinearity (see Appendix D for details). @tbl-coefs shows the summary of the posterior distribution of the fixed regression coefficients, and their degree of overlap with the ROPE. For interpretability, we report the estimated regression coefficients transformed to the probability scale. The resulting values correspond to the maximum difference in probability of acquisition (*Comprehension* or *Comprehension and Production*) that corresponds to a one standard deviation change in each predictor^[The logit and probability scales relate non-linearly. This means that one logit difference is not necessarily translated to a unique value in the probability scale. For example, the probability of acquisition of a given word might increase in 5% when age increases from 22 to 23 months, the probability of acquisition of the same word might only increase in 0.2% when age increases from 31 to 32 months. The linear growth of the probability of acquisition differs along the logistic curve, and therefore deciding the age point at which to report the estimates of the regression coefficients in the probability scale is not trivial. Following @gelman2020regression, we report the maximum value of such coefficient, which corresponds to the linear growth (i.e. derivative) of the logistic curve at the age at which most participants were acquiring a given word. This value can be approximated by dividing the coefficient in the logit scale by four: $\hat{\beta_j}/4$, where $\hat{\beta_j}$ is the estimated mean of the posterior distribution of coefficient $j$.].

```{r tbl-coefs}
#| label: tbl-coefs
#| tbl-cap: "Posterior distribution of regression coefficients."
#| tbl-note: "Median: median of the posterior distribution in the probability scale. 95\\% HDI: 95\\% highest density interval of the distribution; narrowest range of values that contains 95\\% of the distribution, and therefore is 95\\% likely to contain the true value, given the data. On top of each distribution, we indicate the proportion of posterior samples in the 95\\% HDI that fall into the region of practical equivalence (ROPE). This proportion indicates the probability of the true value of the coefficient being equivalent to zero. Lower values indicate that the true value is unlikely to be zero of equivalent to zero."
#| echo: false
#| message: false
#| warning: false
#| out-width: 80%
# summarise posterior draws
intercept_label <- "Intercepts (at 22 months)"

model_summary |>
    select(.variable_name, .median, .lower, .upper, .rope_overlap, .type) |>
    mutate(across(.median:.upper,
                  \(x) ifelse(.type==intercept_label, plogis(x), x/4))) |> 
    gt(groupname_col = ".type", 
       rowname_col = ".variable_name") |>
    fmt_percent(.rope_overlap, decimals = 3) |> 
    fmt_number(c(.median:.upper), decimals = 3) |>
    cols_merge(c(.lower, .upper),
               pattern = "[{1}, {2}]") |>
    cols_label(
        .variable_name = "Parameter",
        .median = "Median",
        .lower = md("95% HDI"),
        .rope_overlap = md("ROPE prob.")
    ) |>
    tab_style(cell_text(weight = "bold"),
              cells_column_labels(columns = 1:5)) |>
    tab_style(cell_text(align = "left"),
              cells_title(groups = "subtitle")) |> 
    tab_style(cell_text(size = "small"), 
              list(cells_body(),
                   cells_row_groups(),
                   cells_stub())) |> 
    tab_style(cell_text(style = "italic"),
              cells_row_groups())
```

@@

```{r fig-marginal}
#| label: fig-marginal
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.5
#| out-width: 100%
#| eval: true
#| fig-cap: "Posterior marginal effects. Thin lines correspond to 50 mean predictions. Thicker lines indicate the median of those predictions. Different colours indicate different levels of cognateness (phonological similarity). Predictions are presented separately for different degrees of word exposure index: little exposure to the word, mean exposure, and high exposure. Predictions for Comprehension are show on top and predictions for Comprehension and Production are shown on the bottom. In-sample predictions lie inside the grey rectangles."
clrs_reorder <- c("#58508d", "#003f5c", "#122A36", "#ffa600", "#ff6361", "#bc5090")

model_epreds |> 
    mutate(age = (age_std * sd(responses$age)) + mean(responses$age),
           exposure = factor(exposure_std,
                             levels = unique(exposure_std),
                             labels = c("Low exposure (-1 SD)",
                                        "Mean exposure",
                                        "High exposure (+1 SD)"),
                             ordered = TRUE),
           lv = factor(lv_std,
                       levels = unique(lv_std),
                       labels = paste0(c(0, 50, 100), "%"),
                       ordered = TRUE),
           .category = ifelse(.category=="Understands",
                              "Comprehension",
                              "Production")) |> 
    ggplot(aes(age, .value, colour = lv, fill = lv)) +
    facet_grid(.category~exposure) +
    annotate(geom = "rect",
             xmin = min(responses$age),
             xmax = max(responses$age),
             ymin = -Inf,
             ymax = Inf,
             fill = "grey95",
             colour = NA) +
    annotate(geom = "text",
             x = mean(responses$age),
             y = 0,
             hjust = -0.1,
             vjust = 0,
             label = "Observed age range",
             size = 2) +
    geom_vline(xintercept = mean(responses$age), 
               linetype = "dotted") +
    geom_hline(yintercept = 0.5,
               linetype = "dotted") +
    geom_line(aes(group = interaction(lv, .draw)),
              linewidth = 0.5,
              alpha = 1/10) +
    stat_summary(fun = median, geom = "line", linewidth = 0.75) +
    labs(x = "Age (months)",
         y = "Probability of acquisition",
         linetype = "Cognateness (phonological similarity)",
         colour = "Cognateness (phonological similarity)",
         fill = "Cognateness (phonological similarity)") +
    guides(colour = guide_legend(override.aes = list(linewidth = 2))) +
    scale_colour_manual(values = clrs_reorder[c(4, 5, 1)]) +
    scale_fill_manual(values = clrs_reorder[c(4, 5, 1)]) +
    scale_y_continuous(breaks = seq(0, 1, 0.25),
                       limits = c(0, 1)) +
    scale_x_continuous(breaks = seq(0, 50, 5)) +
    theme(legend.position = "top",
          legend.key.size = unit(1, "cm"),
          legend.key.height = unit(1, "cm"),
          axis.text.x = element_text(size = 9),
          panel.grid = element_blank(),
          panel.border = element_rect(fill = NA, 
                                      colour = "black", 
                                      linewidth = 0.75))
```

@@

E_e <- bvq_data$pool |>
    mutate(item = gsub("cat_|spa_", "", item),
           xsampa = gsub("\\.", "", xsampa)) |>
    dplyr::filter(te %in% items$te, 
                  item %in% items$item) |>
    select(te, language, xsampa) |> 
    pivot_wider(names_from = language,
                values_from = xsampa, 
                id_cols = te,
                names_repair = janitor::make_clean_names) |>
    dplyr::filter(grepl("E", catalan), 
                  grepl("e", spanish)) |>
    left_join(distinct(items, te, lv),
              by = "te")
```
